---
title: "Amounts and proportions"
subtitle: "Exercise 4 --- PMAP 8551/4551, SEMESTER YEAR"
author: "YOUR NAME HERE"
date: "DATE GOES HERE"
date-format: "long"
format:
  html:
    toc: true
    knitr:
      opts_chunk: 
        dev: "ragg_png"
        dpi: 300
  typst:
    toc: true
    toc-title: Contents
    output-file: PREFIX_04-exercise.pdf
    pdf-standard: ua-1
    callout-appearance: minimal
    include-in-header: 
      text: |
        // TOC stuff
        // Make the title centered and slightly smaller than a full heading
        #show outline: it => {
          show heading: it => align(center, text(size: 0.75em, it))
          it
        }
    include-before-body: 
      text: |
        // Color links with purple from viridis plasma: 
        // viridisLite::viridis(8, option = "plasma", begin = 0.1, end = 0.9)[3]
        #show link: set text(fill: rgb("9a169f"))

        // Keep track of previous heading levels
        #let last-heading-level = state("last-heading-level", 0)

        // Add page breaks before all level 1 headings
        // Add page breaks before level 2 headings that don't come after level 1
        #show heading: it => {
          context {
            let prev-level = last-heading-level.get()

            if it.level == 1 {
              pagebreak(weak: true)
              last-heading-level.update(1)
            } else if it.level == 2 {
              if prev-level != 1 {
                pagebreak(weak: true)
              }
              last-heading-level.update(2)
            } else {
              last-heading-level.update(it.level)
            }
          }
          it
        }
  pdf:
    toc: true
  docx:
    toc: true
---

# Task 1: Session check-in

::: {.callout-tip title="Answer"}
Hopefully you wrote some interesting and muddy things!
:::


# Task 2: Long Beach Animal Shelter

```{r}
#| label: setup
#| warning: false
#| message: false

# Load libraries here
library(tidyverse)
```

## Data description and cleaning

This data comes from the [City of Long Beach Animal Care Services](https://www.longbeach.gov/acs/) via the [{animalshelter}](https://emilhvitfeldt.github.io/animalshelter/) R package (and it was a [#TidyTuesday dataset in March 2025](https://github.com/rfordatascience/tidytuesday/blob/main/data/2025/2025-03-04/readme.md)).

There are 22 columns, and [they're documented here](https://github.com/rfordatascience/tidytuesday/blob/main/data/2025/2025-03-04/readme.md#data-dictionary)---make sure you consult that page!

The code below loads and cleans the data, mostly by filtering it to include the most common types of animals, intake types, and outcome types. **You don't need to modify anything in this chunk of code---you only need to run it.**

```{r}
#| label: load-clean-data
#| warning: false
#| message: false

# Load the full data
animals_raw <- read_csv("data/animals.csv")

animals <- animals_raw |> 
  # Only keep the most common animals, intake types, and outcome types
  filter(animal_type %in% c("bird", "cat", "dog", "wild", "other")) |>
  filter(intake_type %in% c("owner surrender", "stray", "wildlife")) |>
  filter(
    outcome_type %in%
      c("outcome_type", "rescue", "adoption", "euthanasia", "transfer",
        "return to owner", "shelter, neuter, return", "died",
        "community cat", "return to wild habitat")
  ) |>
  # Add columns for the year and month
  mutate(
    intake_year = year(intake_date),
    intake_month = month(intake_date, label = TRUE, abbr = FALSE)
  ) |> 
  mutate(
    across(c(animal_type, intake_type, outcome_type), 
    \(x) str_to_title(x))
  )
```


## Animal type

What are the most common type of animals seen by the shelter?

### Recreate data summary

Right now there's a row for each animal visit. We need to condense that down to get counts of different types of variables. We can do this by using `group_by()` and `summarize()`.

Use `group_by()` and `summarize()` to create a data frame named `animal_counts` that looks like this (hint---in `summarize()`, you'll want to use the `n()` function to calculate the number of rows in each group):

```{r}
#| label: recreate-me-data-1
#| echo: false
#| ref.label: recreation-data-1
```

::: {.callout-tip title="Answer"}

```{r}
#| label: recreation-data-1

animal_counts <- animals |> 
  group_by(animal_type) |> 
  summarize(total = n())
animal_counts
```

:::

### Recreate bar plot

Run the code chunk below to see a plot.

```{r fig.path="images/"}
#| label: recreate-me-1
#| echo: false
#| ref.label: recreation-plot-1
#| opts.label: true
```

Use `ggplot()` and `geom_col()` in the chunk below to re-create the plot above using the `animal_counts` dataset. (*note*---the bars are filled, but there's no legend)

::: {.callout-tip title="Answer"}

```{r}
#| label: recreation-plot-1
#| fig-alt: |
#|   Bar chart of animal counts by type at the Long Beach Animal Shelter.
#|   Animal types (Bird, Cat, Dog, Other, Wild) on the x-axis, total visit
#|   count on the y-axis, each bar filled with a distinct color but no legend.
#|   Dogs and cats account for the large majority of shelter animals, while
#|   birds and wild animals are far less frequent.

ggplot(animal_counts, aes(x = animal_type, y = total, fill = animal_type)) +
  geom_col() +
  guides(fill = "none")
```

:::

## Outcomes

What are the most common types of outcomes for animals taken to the shelter? What proportion are adopted or rescued?

### Recreate data summary

Use `group_by()` and `summarize()` and `mutate()` to create a data frame named `outcome_props` that looks like this (hint---that proportion column is calculated like `total / sum(total)`):

```{r}
#| label: recreate-me-data-2
#| echo: false
#| message: false
#| ref.label: recreation-data-2
```

::: {.callout-tip title="Answer"}

```{r}
#| label: recreation-data-2
#| message: false

outcome_props <- animals |> 
  group_by(outcome_type) |> 
  summarize(total = n()) |> 
  mutate(proportion = total / sum(total))
outcome_props
```

:::

### Recreate lollipop plot

Run the code chunk below to see a plot.

```{r fig.path="images/"}
#| label: recreate-me-2
#| echo: false
#| ref.label: recreation-plot-2
#| message: false
#| opts.label: true
```

Use `ggplot()` and `geom_pointrange()` in the chunk below to re-create the plot above using the `outcome_props` dataset.

::: {.callout-tip title="Answer"}

```{r}
#| label: recreation-plot-2
#| message: false
#| fig-alt: |
#|   Lollipop chart of shelter animal outcome proportions. Outcome types on
#|   the y-axis in alphabetical order, proportion of all outcomes on the
#|   x-axis. Transfer, rescue, euthanasia and adoption are among the most common 
#|   outcomes; died and return to wild habitat are the least frequent. The
#|   alphabetical ordering makes comparing across types difficult.

ggplot(outcome_props, aes(x = proportion, y = outcome_type)) +
  geom_pointrange(aes(xmin = 0, xmax = proportion))
```

:::

Currently the outcome types are sorted alphabetically, which isn't super helpful. Sort them by proportion and recreate the plot below.

```{r fig.path="images/"}
#| label: recreate-me-3
#| echo: false
#| ref.label: recreation-plot-3
#| message: false
#| opts.label: true
```

*Big hint!* Add this to the end of the code to make `outcome_props` to order `outcome_type` by the proportion column:

```r
CODE_FOR_OUTCOME_PROPS |> 
  mutate(outcome_type = fct_reorder(outcome_type, proportion))
```

::: {.callout-tip title="Answer"}

```{r}
#| label: recreation-plot-3
#| message: false
#| fig-alt: |
#|   Lollipop chart of shelter animal outcome proportions, sorted from least
#|   to most common. Outcome types on the y-axis ordered by ascending
#|   proportion, proportion of all outcomes on the x-axis. Sorting by
#|   proportion makes it immediately clear which outcomes are most and least
#|   frequent.

outcome_props_sorted <- outcome_props |> 
  mutate(outcome_type = fct_reorder(outcome_type, proportion))

ggplot(outcome_props_sorted, aes(x = proportion, y = outcome_type)) +
  geom_pointrange(aes(xmin = 0, xmax = proportion))
```

:::

## Intake type over time

Are there seasonal patterns in the types of animal intake at the shelter?

### Recreate data summary

Use `group_by()` and `summarize()` and `mutate()` to create a data frame named `month_intake_props` that looks like the output below.

You'll need to group by two variables to make the summary. IMPORTANTLY you'll also need to add another `group_by()` in between `summarize()` and `mutate()`, otherwise, R will calculate percentages in unexpected groups.

If you want the percentages of months to add up to 100% in each intake type (i.e. the percent of adoptions in January, February, March, etc.), you'll want to group by `intake_type` before calculating the proportion; if you want the percentages of intake types to add up to 100% in each month (i.e. the percent of all different intake types in January), you'll want to group by `intake_month` before calculating the proportion.

```{r}
#| label: recreate-me-data-4
#| echo: false
#| message: false
#| ref.label: recreation-data-4
```

::: {.callout-tip title="Answer"}

```{r}
#| label: recreation-data-4
#| message: false

month_intake_props <- animals |> 
  group_by(intake_type, intake_month) |> 
  summarize(total = n()) |> 
  group_by(intake_type) |> 
  mutate(proportion = total / sum(total))
month_intake_props
```

:::

### Recreate heatmap

Run the code chunk below to see a plot.

```{r fig.path="images/"}
#| label: recreate-me-4
#| echo: false
#| ref.label: recreation-plot-4
#| message: false
#| opts.label: true
```

Use `ggplot()` and `geom_tile()` in the chunk below to re-create the plot above using the `outcome_props` dataset.

*Hint*: By default, the months will appear reversed, with January at the bottom and December at the top. Use `y = fct_rev(intake_month)` to reverse them.

::: {.callout-tip title="Answer"}

```{r}
#| label: recreation-plot-4
#| message: false
#| fig-alt: |
#|   Tile heatmap of animal intake proportions by month and intake type.
#|   Three intake types (Owner Surrender, Stray, Wildlife) on the x-axis,
#|   months (January at top through December at bottom) on the y-axis. Color
#|   encodes the proportion of annual intake occurring in each month for each
#|   intake type. Stray and wildlife intakes concentrate in spring and summer;
#|   owner surrenders are more evenly spread throughout the year.

ggplot(
  month_intake_props,
  aes(x = intake_type, y = fct_rev(intake_month), fill = proportion)
) +
  geom_tile()
```

:::

# Task 3: Extension

Copy the code for one of your plots above and paste it into the chunk below. Do some extra things to it, like changing the labels, changing the colors, adding a new geom, plotting a different variable from the dataset (there are lots of them!), changing the theme, or whatever. **This is your chance to play around and explore.**

::: {.callout-tip title="Answer"}

Hopefully you did something neat! Here's something I did with the heatmap:

```{r}
#| label: extension
#| message: false
#| warning: false
#| fig-alt: |
#|   Tile heatmap of animal intake proportions by month and intake type,
#|   styled with the plasma color scale and percent-formatted legend. Three
#|   intake types on the x-axis, months (January at top) on the y-axis.
#|   Bright yellow-orange cells highlight that wildlife and stray intakes
#|   spike strongly in spring and summer, supporting the title "Wildlife and
#|   strays most common in spring and summer."

library(scales)  # For better labels, like label_percent()

ggplot(
  month_intake_props,
  aes(x = intake_type, y = fct_rev(intake_month), fill = proportion)
) +
  geom_tile() +
  scale_fill_viridis_c(option = "plasma", labels = label_percent()) +
  labs(
    x = "Intake type",
    y = NULL,
    fill = "Proportion per month\nfor each intake type",
    title = "Wildlife and strays most common\nin spring and summer"
  ) +
  theme_minimal()
```

:::
